### 2021-08-26 plos_latex_template.tex
RJ: version 2 of the conclusion

%ast: matt's original sentence: reaching users and developers of the R language from more than 120 countries.

@@

%ast will write a paragraph about why diversity and the systemic inequalities
While there has been some action to address systematic inequalities, there is still a lot of room for improvement.

@@

\section{Rule X: Define what diversity and inclusion mean for your conference}
\label{rule_diversity}

Each one of us can have different points of view about the meaning of diversity, depending on where we come from. If you are reading this, you may be here because you already have some experience in Diversity, Inclusion, and Accessibility practices. Or you may want to improve diversity in your conference but do not know how or where to start. This article does not intend to be a comprehensive review of this topic --this would be impossible-- but will give some ideas pointing in key directions.    
% RJ: Most of the first paragraph could pivot to the intro, but we can see later.

@@

To accomplish this diversity vision, you will have to take conscious decisions toward removing bias from your conference's spaces. Having diverse people in decision-making positions will affect positively all the other aspects of your conference because all the decision-making processes will benefit from their input (see Rule \ref{rule_organizing_team}), and because representation matters (see Rule \ref{rule_communication}).
% RJ: we could improve the wording of the last sentence.

\section{Create a safe and welcoming environment}

@@

%Representation is an important aspect of diversity and inclusion: seeing people like you, from your country, that speak your language is one of the best way to feel you belong

@@

Paying attention to some dimensions of diversity while neglecting others may have unintended consequences. Lack of representation, unwelcoming --or overtly aggressive-- environments hinder participation (or future participation) of people who could otherwise become active community members. In severe cases, they can divert career paths, affect lives, and exclude people from some fields.

%So take this seriously, think about the reasons why you want to create a more diverse conference, and be open to learn. Do not treat diversity as a checklist, do not use people from minoritized groups for image purposes (tokenization), and do not act as a savior but as a genuine collaborator. %ast: some version of this might come in a box.

A key component in creating a safe environment during a conference is adopting a code of conduct and creating and training a team to enforce it \cite{favaroYourScienceConference2016}. The code of conduct is a document meant to keep the community safe and should state clearly: the unacceptable behaviors, the spaces of the conference in which it applies, the consequences for engaging in unacceptable behavior, and the way to report violations \cite{auroraHowRespondCode2018}. 
The code of conduct should be displayed prominently in several spaces of the conference to deter people from unacceptable behavior. An efficient code of conduct acts as a protection for the community because the people who may be target of unacceptable behavior tend to be people with less power or privilege.

\section{Rule X: Have a strong online component of the conference} 
\label{rule_online}
In-person interaction is priceless; however, it is more expensive for some, even unattainable for others. This inaccessibility is particularly true for global conferences that usually take place in high-income countries, making it financially demanding for international participants and often impossible to attend due to immigration requirements \cite{arendDisparityConferenceRegistration2019,gewinWhatScientistsShould2019}. Online conferences are more inclusive: they do not need a visa or a big budget, and are more accessible to people who may be unable to travel because of health issues or family responsibilities. This means that online conferences have a greater reach, not only in terms of participants but in terms of the tutors and presenters that can participate \cite{atkinsonJournalMedicine20202021}. Furthermore, an online format is more environmentally friendly since it eliminates travel-related emissions \cite{sarabipourChangingScientificMeetings2021,ninerBetterWhomLeveling2021, gattrellComparisonCarbonCosts2021}.

@@

\section{Rule X: Have an inclusive and diverse organizing team}
\label{rule_organizing_team}
A genuinely inclusive conference can only be organized by an inclusive and diverse organizing team. Build a team with people from different regions, genders, socioeconomic statuses, and other aspects of diversity. Particular attention should be paid to the usually marginalized groups (see \textbf{Rule \ref{rule_diversity}} discussing the dimensions of diversity lacking in your community and event). To ensure a deeper understanding and smooth communication with different diverse groups, it is essential to create a representative working group that functions as a snapshot of the community at large. But this will only work if there is real inclusion. One would be surprised by how often conference leads tend to do everything themselves instead of reaching out to the right people for the right task -- likely because the latter involves careful planning and communication well ahead of time or because they naively overestimate their expertise in areas they lack practical experience in. For example, a cisgender or abled person estimating what might work best for transgender people or people with sensory disabilities. The same holds for other dimensions of diversity. A truly inclusive and welcoming space is one in which everyone in the team is invited and allowed to bring their experience to bear. Even if creating and maintaining such a team and space is challenging, the positive outcomes far outweigh any minor inconveniences (e.g. time zones), and the team will grow, learn, and embrace more inclusive practices \cite{hongGroupsDiverseProblem2004}.

%The team may also have to review carefully and teach themselves the correct vocabulary for internal and external communication and identify the best ways to account for every culture and situation. % rj: left this out, Heather said it doesn't make much sense here %ast: we can collect part of this in rule 1, since some self-teaching, and not expecting to be taught, is necessary. nope. now i think this is here.

%%Disabled people often repeat "Nothing about us without us", and this is true for every marginalized group. This means the actual life experiences, expertise, and insights from people in marginalized groups are not replaceable with good intentions from people outside these groups.

@@

\section{Rule X: Make the conference accessible to people with disabilities}
\label{rule_accessibility}

Conferences are among the least accessible spaces that people with disabilities may encounter \cite{priceAccessImaginedConstruction2009}.
%People may also find conferences inaccessible due to their caring responsibilities, dietary restrictions, religious practices, or LGBTQ+ status.
Ironically, accessibility practices are inclusive not only for people with disabilities but can be beneficial to a broad spectrum of people. For instance, having captions is helpful to deaf and hard-of-hearing people, non-native speakers, and everyone in general.

@@

Most importantly, accessibility practices are not afterthoughts that can be dealt with at the last minute. 
They require time and early decision-making \cite{irishIncreasingParticipationUsing2020}. 
Conversely, inaccessible decisions are hard to course-correct, e.g., when finding out too late that a venue is inaccessible for wheelchairs in an in-person conference. 
Welcoming people that are usually disabled by standard practices in conferences into the organizing team would allow them to take part in the decisions from the beginning and spot right away the inaccessible practices that need to improve (see \textbf{Rule \ref{rule_organizing_team}}).

\section{Rule X: Consciously unbias your spotlight roles}
\label{rule_spotlight}

@@

\section{Rule X: Set fair registration rates and provide financial support when needed}

@@

\section{Rule X: Don't let language restrict high-quality participation}
\label{rule_language}

@@

\section{Rule X: Actively reach out and involve people who have been systematically excluded}
\label{rule_communication}
%this rule name should be more explicit towards communication, "promoting"
If part of the community has been historically discriminated against, one should emphasize that they are particularly welcome in this event and that the organization will make it a safe and inclusive environment. Publish the diversity statement (e.g., \url{https://user2021.r-project.org/about/diversity_statement/}), the code of conduct and accessibility guidelines. Fee waivers, financial support, and ease of application/registration should also be advertised; people who fear getting rejected when applying for scholarships and waivers may find it relieving that the process will be supportive rather than discriminatory against underrepresented groups. Advertise the conference in multiple languages (rule \ref{rule_language}). Express this welcoming spirit in your communication strategy (social networks, website, brand, and visual identity) to let people know that they are seen, respected, and welcome; that this is their space and community too.

% RJ: could add some ideas from here 
% https://www.science.org.au/files/userfiles/support/emcr/documents/one-page-summary-emcr-improving-diversity-web.pdf

\section{Rule X: Do not expect the organizing team to work full time and for free}

@@

\section{Rule X: Don't let perfect be the enemy of good}
 While conference teams should have a bold vision and clear targets consensus about where to shine and where to compromise is equally important. As a team, this is a challenging journey. It means letting go of some ideals, or committing time and money into something that one does not find particularly useful or important.

Remember, aspirations can be local and narrow and as a consequence shy away valuable contributors. A late bus is almost certainly better than no bus. A broader pool of contributors does not only allow organizers to add more and work on multiple aspects at a different pace, it also helps to split workload and responsibilities to avoid burning out over the marathon that organizing a conference is.

@@

@@

The ten rules stated here can be adapted depending on the conference format and settings. From our own experience, %it's worth it :P
we are aware that enforcing these rules requires a tremendous amount of work. We engaged in most of these practices, and learned others too late, so we share them here as part of our learning process. 
Since accessibility and inclusiveness are not as widespread as they should be, the process might seem exacting. But this is where the power of communities and individuals reside. If more conferences and domains employ these techniques, the process will get more streamlined, straightforward, and mainstream to adapt with minimal overhead. And we will make a change towards healthier, stronger, and more inclusive communities, empowering historically marginalized people and learning from them, thereby increasing the overall quality of our conferences.%After all, the strength of any community lies in its ability to be adaptive, inclusive, and accessible.

>>>>

%andrea: matt's original sentence: reaching users and developers of the R language from more than 120 countries.

@@

%andrea will write a paragraph about why diversity and the systemic inequalities
While there has been some action to address systematic inequalities, there is still a lot of room for improvement.
% Rocío: Somehow this part below should be linked to the systemic inequalities part
Each one of us can have different points of view and experiences regarding inequality, depending on where we come from.
If you are reading this, you may be here because you already have some experience in Diversity, Inclusion, and Accessibility practices. Or you may want to improve diversity in your conference but do not know how or where to start. This article does not intend to be a comprehensive review of this topic --this would be impossible-- but will give some ideas pointing in key directions.   %Rocío: And finish with something more specific to connect with the next paragraph.

@@

\section{Define what diversity and inclusion mean for your conference}
\label{rule_diversity}

@@

\section{Create a safe and welcoming environment}
\label{rule_inclusion}

@@

Paying attention to some dimensions of diversity while neglecting others may have unintended consequences. Lack of representation, unwelcoming --or overtly aggressive-- environments hinder participation (or future participation) of people who could otherwise become active community members. In severe cases, they can divert career paths, affect lives, and exclude people from some fields. %Rocío: it would be nice to have reference here.

Adopting a code of conduct and creating a team to enforce it are key aspects in creating a safe environment during a conference \cite{favaroYourScienceConference2016}. The code of conduct is a document meant to keep the community safe and should state clearly: the unacceptable behaviors, the spaces of the conference in which it applies, the consequences for engaging in unacceptable behavior, and the way to report violations \cite{auroraHowRespondCode2018}. 
The code of conduct should be displayed prominently in several spaces of the conference to deter people from unacceptable behavior. An efficient code of conduct acts as a protection for the community because the people who are the target of unacceptable behavior tend to be the ones with less power or privilege.

%So take this seriously, think about the reasons why you want to create a more diverse conference, and be open to learn. Do not treat diversity as a checklist, do not use people from minoritized groups for image purposes (tokenization), and do not act as a savior but as a genuine collaborator. %andrea: some version of this might come in a box.

@@

\section{Have a strong online component of the conference} 
\label{rule_online}
In-person interaction is priceless; however, it is more expensive for some, and even unattainable for others. This inaccessibility is particularly true for global conferences that usually take place in high-income countries, making it financially demanding for international participants and often impossible to attend due to immigration requirements \cite{arendDisparityConferenceRegistration2019,gewinWhatScientistsShould2019}. Online conferences are more inclusive: they do not need a visa or a big budget, and are more accessible to people who may be unable to travel because of health issues or family responsibilities. This means that online conferences have a greater reach, not only in terms of participants but in terms of the tutors and presenters that can participate \cite{atkinsonJournalMedicine20202021}. Furthermore, an online format is more environmentally friendly since it eliminates travel-related emissions \cite{sarabipourChangingScientificMeetings2021,ninerBetterWhomLeveling2021, gattrellComparisonCarbonCosts2021}.

@@

\section{Have an inclusive and diverse organizing team}
\label{rule_organizing_team}

A genuinely inclusive conference can only be organized by an inclusive and diverse organizing team. Build a team with people from different regions, genders, socioeconomic statuses, and other aspects of diversity. Particular attention should be paid to the usually marginalized groups (see \textbf{Rule \ref{rule_diversity}} discussing the dimensions of diversity lacking in your community and event). To ensure a deeper understanding and smooth communication with different diverse groups, it is essential to create a representative working group that functions as a snapshot of the community at large. If you already have an organizing team, check for biases in its composition.

%liz: please check if this paragraph connects well with the whole rule. i am using nothing about us without us but also talking about all marginalized groups, in the context of building a team. att: andrea

Gathering a diverse team will only work if there is real inclusion. Disabled people often say: "Nothing about us without us"; the same holds for other dimensions of diversity. This means that the actual life experiences, expertise, and insights from people in marginalized groups are not replaceable with good intentions from people outside these groups. A truly inclusive and welcoming space is one in which everyone in the team is invited and allowed to bring their experience to bear. So, if you plan to organize something related to any community, region, or specific dimension of diversity, involve people from these groups. Having diverse people in decision-making positions will affect positively all the other aspects of your conference because all the processes will benefit from their input, expertise, and distinct perspectives \cit{hong_groups_2004}. On the other hand, seeing diverse teams is important to create a welcoming space because seeing people with similar life experiences occupy public spaces, positions of power, and breaking negative stereotypes (i.e., representation) matters (see Rule \ref{rule_inclusion}). %Representation is an important aspect of diversity and inclusion: seeing people like you, from your country, that speak your language is one of the best way to feel you belong

Creating and maintaining such a team and space may seem more challenging than working in homogeneous teams, but the positive outcomes far outweigh any inconveniences and the team will grow, learn, and embrace more inclusive practices \cite{hongGroupsDiverseProblem2004}.

%How to work: 
Most advice when working with a diverse team should apply to every team, but is especially true when working with people from marginalized groups. First, delegate. One would be surprised by how often conference leads tend to do everything themselves instead of reaching out to the right people for the right task -- likely because the latter involves careful planning and communication well ahead of time or because they naively overestimate their expertise in areas they lack practical experience in.

Don't expect self-nomination and voting to work as mechanisms to counteract systemic inequalities. Nominate directly and offer the positions of power, even if this means stepping down. Offer mentorship and guidance if you encounter doubts, cases of impostor syndrome, doubts about the use of English in communication \ref{rule_language}. Break the expectations about lonley leadership and create smaller, co-led groups.

In addition to exclude derogatory or discriminatory language (which would constitute a violation of the code of conduct \ref{rule_inclusion} but also, should never happen), make the effort to teach yourself the best ways to communicate to account for every culture and situation.

Finally, splitting the workload and responsibilities should not be done by putting care-taking labors --community building, meeting organization and note taking, conversations with potential partners-- on the hands of women and other minoritized groups .

Take care of your team!

@@

\section{Make the conference accessible to people with disabilities}
\label{rule_accessibility}

Conferences are among the least accessible spaces that people with disabilities may encounter \cite{priceAccessImaginedConstruction2009}.
Ironically, accessibility practices are inclusive not only for people with disabilities but can be beneficial to a broad spectrum of people. For instance, having captions is helpful to deaf and hard-of-hearing people, non-native speakers, and everyone in general.

@@

Most importantly, accessibility practices are not afterthoughts that can be dealt with at the last minute. 
They require time and early decision-making \cite{irishIncreasingParticipationUsing2020}. 
Conversely, inaccessible decisions are hard to course-correct, e.g., when finding out too late that a venue is inaccessible for wheelchairs in an in-person conference. 
On the other hand, it is likely that financial constraints may not allow you to make your conference completely accessible and there will be choices to make (e.g. choosing between captions and interpreters). 
Welcoming people that are usually disabled by standard practices in conferences into the organizing team would allow them to take part in the decisions from the beginning, spot right away the inaccessible practices that need to improve (see \textbf{Rule \ref{rule_organizing_team}}), and help make the right choices.

\section{Consciously unbias your spotlight roles}
\label{rule_spotlight}

@@

\section{Set fair registration rates and provide financial support when needed}

@@

\section{Don't let language restrict high-quality participation}
\label{rule_language}

@@

\section{Actively reach out to people who have been systematically excluded}
\label{rule_communication}
%Andrea: this rule name should be more explicit towards communication, "promoting"
If part of the community has been historically discriminated against, one should emphasize that they are particularly welcome in this event and that the organization will make it a safe and inclusive environment. Publish the diversity statement (e.g., \url{https://user2021.r-project.org/about/diversity_statement/}), the code of conduct and accessibility guidelines. Fee waivers, financial support, and ease of application/registration should also be advertised; people who fear getting rejected when applying for scholarships and waivers may find it relieving that the process will be supportive rather than discriminatory against underrepresented groups. Advertise the conference in multiple languages (rule \ref{rule_language}). Express this welcoming spirit in your communication strategy (social networks, website, brand, and visual identity) to let people know that they are seen, respected, and welcome; that this is their space and community too. 
% Yani: We need to work a little more here, perhaps with some examples.
% Rocío: could add some ideas from here 
% https://www.science.org.au/files/userfiles/support/emcr/documents/one-page-summary-emcr-improving-diversity-web.pdf

\section{Do not expect the organizing team to work full time and for free}

@@

\section{Don't let perfect be the enemy of good}
 While conference teams should have a bold vision and clear targets consensus about where to shine and where to compromise is equally important. As a team, this is a challenging journey. It means letting go of some ideals, or committing time and money into something that one does not find particularly useful or important.
%Yani: ughhhh... this doesn't sound good. I agree more with let go some ideal, for example, when I compromise in not have caption for tutorials, because we don't have money. Have caption in the other part of the conf was better because reach a more wide audience of the conference. I will try to show the compromise from the point of view of all the things we know we still have to do about inclusion, but we cannot do because we don't have money or the knowledge or the capacity at this time. This is also mention in some of the other rules. I don't remember if it is the one with the goals and indicators.

Remember, aspirations can be local and narrow and as a consequence shy away valuable contributors. A late bus is almost certainly better than no bus. 
% Yani: I don't like this phrase. It is like we have something of a lower quality than the original idea, when I truly think we have a high quality conference because was inclusive and diverse. and science say that a diverse team get better results :)
A broader pool of contributors does not only allow organizers to add more and work on multiple aspects at a different pace, it also helps to split workload and responsibilities to avoid burning out over the marathon that organizing a conference is.
%Yani: We have a broader team because we want a more diverse pool of view and experience, this is a good phrase but sound like we people were an object. Perhaps we can rephrase and add to the rule that talk about the composition of the team.%andrea: i am taking split workload and responsibilities up, but they were already there.

@@

%\section{Rule X: Have a bold vision but be ready to compromise when necessary}
This rule is a complement to all the other rules. It is important to have a bold vision and clear targets. Some parts might be defined as baseline assumptions that are not up for discussion. However, there will always be compromises. Ideal solutions that meet all our expectations are often not possible. Don't let the perfect be the enemy of good. As a team, this is a challenging journey. It means letting go of some ideals, or committing time and money into something that one does not find particularly useful or important. We organized useR! during a global pandemic. Very early on, we tried to recognize that our resources were limited (both in time of the organizing team but also in money). We had to let go of certain aspects that were important to many of us. For an inclusive conference, we need to avoid giving in because considering all aspects seems impossible and burning ourselves and the team out because considering all aspects is actually impossible.

@@

Since accessibility and inclusiveness are not as widespread as they should be, the process might seem exacting.
But this is where the power of communities and individuals reside. 
If more conferences and domains apply these rules, the process will get more streamlined, straightforward, and mainstream to adapt with minimal overhead. 
And you will make a change towards healthier, stronger, and more inclusive communities, empowering historically marginalized people and learning together, thereby increasing the overall quality of your conferences.

We engaged in most of the practices mentioned here, and learned others along the way, so we share them here as part of our learning process. 
We organized useR! during a global pandemic  
and, as a team, this was a challenging journey. 
Very early on, we tried to recognize that our resources were limited (both in time of the organizing team and financially). 
We had to let go of certain aspects that were important to many of us. From that experience, this is our last message: 
Your conference may not become perfectly inclusive and accessible. 
Don't let perfect be the enemy of good.

% Try not to burn you and the team out trying to make it perfect.
